---
title: "first_try_project"
author: "Felix"
format: html
editor: visual
---

# general points:

-   hosp_patients can be deleted as well as hosp_patients_per_million
-   

## try to answer questions: \## Question 1: How did vaccination coverage in Germany develop during the COVID-19 pandemic?

**Claim type:** Descriptive

We aim to describe how vaccination coverage changed between 2020 and 2024. In particular, we want to identify important phases of the vaccination campaign and examine how quickly vaccination rates increased over time.

Mostly it´s better to work with the smoothed data columns. Argue why!

Something like: There may be∂days, like sunday or holidays, when no new cases/deaths/ numbers in general got added to any statistics. IN the smoothed version these days do not exist because in the mean over more days these days don´t matter. Aso new cases have only been added to our statistics sundays, so on every other day we have 0 new cases, what is obvious not true, so with the smoothed verison we get at least the mean for evrey day per week, what is exactly enough for our project.

At first it is interesting, when the first person was vaccinated.

```{r}
library(tidyverse)
```

Now we can only look at the time, in which people got vaccinated. Here we can create a lineplot with new-vaccinated_smoothed on y axis.

```{r}
covid_germany %>% ggplot(aes(x = date, y = new_vaccinations_smoothed))+geom_line()

```

Also while taking a look at new_cases_smoothed or new_death_smoothed

```{r}
covid_germany %>% ggplot(aes(x = date))+
                           geom_line(aes(y = new_vaccinations_smoothed))+
                           geom_line(aes(y = new_cases_smoothed))


```

```{r}
covid_germany %>% ggplot(aes(x = date, 
                             y = hosp_patients))+
  geom_line()

```

To address this question, we will analyse the following variables:

-   `date`
-   `new_vaccinations(_smoothed)`,
-   `people_fully_vaccinated(_smoothed)_per_hundred`

and similar ones.

## möglich für question 2a

```{r}
install.packages("infer")
```

```{r}
library(tidyverse)
library(infer)

# Prepare data for the test:
# We compare mortality between days with lower and higher vaccination coverage.
infer_data <- covid_germany |>
  filter(
    !is.na(people_vaccinated_per_hundred),
    !is.na(new_deaths_smoothed)
  ) |>
  mutate(
    vaccination_group = if_else(
      people_vaccinated_per_hundred >= median(people_vaccinated_per_hundred),
      "high vaccination",
      "low vaccination"
    )
  )

# Observed test statistic:
# Difference in mean smoothed deaths between high- and low-vaccination periods.
obs_stat <- infer_data |>
  specify(new_deaths_smoothed ~ vaccination_group) |>
  calculate(
    stat = "diff in means",
    order = c("high vaccination", "low vaccination")
  )

# Null distribution:
# Under the null hypothesis, mortality is not associated with vaccination group.
null_dist <- infer_data |>
  specify(new_deaths_smoothed ~ vaccination_group) |>
  assume("t")

# p-value:
# Probability of observing a difference at least as extreme as ours under the null.
p_value <- null_dist |>
  get_p_value(
    obs_stat = obs_stat,
    direction = "two-sided"
  )

# Confidence interval:
# 95% confidence interval for the difference in mean mortality.
conf_int <- null_dist |>
  get_confidence_interval(
    point_estimate = obs_stat,
    level = 0.95
  )

# Show results
obs_stat
p_value
conf_int
```

Du kannst direkt unter deinem `{infer}`-Output etwa Folgendes schreiben:

> **The observed difference in means is -82.58, indicating that the average number of smoothed daily COVID-19 deaths was about 83 lower during periods with higher vaccination coverage than during periods with lower vaccination coverage. The p-value is effectively zero (p \< 0.001), providing strong evidence that the observed difference is unlikely to have occurred by chance under the null hypothesis of equal means. The 95% confidence interval ranges from -103.49 to -61.68, suggesting that the true difference in mean daily deaths is likely between approximately 62 and 103 fewer deaths during periods with higher vaccination coverage. Since this is an exploratory analysis based on observational data, these results indicate an association rather than a causal effect of vaccination.**

Oder, falls ihr es etwas kürzer halten wollt:

> **The average number of smoothed daily COVID-19 deaths was significantly lower during periods with higher vaccination coverage (difference in means = -82.58, 95% CI \[-103.49, -61.68\], p \< 0.001). This suggests a strong association between vaccination coverage and mortality, but due to the observational nature of the data, no causal conclusions can be drawn.**

Die zweite Version würde ich für das Proposal bevorzugen, da sie prägnant ist und gleichzeitig die statistischen Ergebnisse korrekt interpretiert.
