# ***Proposal for StatProg2 Projekt***

#### by Lukas Seibold and Felix Obermaier

## Dataset and research questions:

-   **Source and licence:**

    -   The Data set is from the following website: [OWID - Covid Dataset](https://ourworldindata.org/explorers/covid?Metric=Excess+mortality+%28estimates%29&Interval=Cumulative&Relative+to+population=true&country=USA~BRA~JPN~DEU)

    -   It can be downloaded in the following github repository: [OWID - github repository COVID](https://github.com/owid/covid-19-data/tree/master/public/data)

    -   The dataset is called owid-covid-data.csv

    -   The Dataset contains all Data on Covid in the years 2020 - 2024 in all countries of the world

    -   We decided we only want to look at german Data on covid so we filtered it by country

```{r}
library(readr)
library(tidyverse)

covid_germany = read_delim("data/covid_germany.csv")

```

-   **Licence:**

    -   The *Our World in Data Covid Dataset* combines COVID-19 data from several official and scientific sources, including:

        -   [Johns Hopkins University (JHU CSSE)](https://github.com/CSSEGISandData/COVID-19?utm_source=chatgpt.com) for confirmed cases and deaths

        -   official national health authorities for vaccination and hospitalization data

        -   the [Oxford COVID‑19 Government Response Tracker](https://www.bsg.ox.ac.uk/research/research-projects/covid-19-government-response-tracker?utm_source=chatgpt.com) for policy measures

        -   international organizations such as the UN, World Bank, OECD, and IHME for demographic and socioeconomic indicators.

        The dataset is published under the **Creative Commons Attribution (CC BY) license**.

        This license allows users to:

        -   use the data freely,

        -   copy and redistribute it,

        -   modify and adapt it,

        -   use it commercially,

        -   and include it in research or presentations,

        as long as proper attribution to Our World in Data and the original data sources is provided.

## What is in the data set?

The data set contains daily COVID-19 observations for Germany from January 2020 until August 2024.

-   Number of rows: 1674
-   Number of columns: 67

Each row represents one day of reported COVID-19 data in Germany.

The data set includes information on:

-   COVID-19 cases (e.g. `total_cases`, `new_cases`, `new_cases_smoothed`)
-   COVID-19 deaths (e.g. `total_deaths`, `new_deaths`)
-   Reproduction rate (`reproduction_rate`)
-   Hospital and ICU patients (`hosp_patients`, `icu_patients`)
-   Vaccination indicators
-   Testing statistics
-   Socioeconomic indicators and demographic variables

The key variables we expect to use are:

-   `date`
-   `new_cases`
-   `new_cases_smoothed`
-   `total_cases`
-   `new_deaths`
-   `total_deaths`
-   `reproduction_rate`

## Research Questions

### Question 1: How did vaccination coverage in Germany develop during the COVID-19 pandemic?

**Claim type:** Descriptive

We aim to describe how vaccination coverage changed between 2020 and 2024. In particular, we want to identify important phases of the vaccination campaign and examine how quickly vaccination rates increased over time.

------------------------------------------------------------------------

### Question 2: Are higher vaccination rates associated with lower COVID-19 mortality in Germany?

**Claim type:** Exploratory

We investigate whether periods with higher vaccination coverage tend to be associated with lower COVID-19 death rates. The goal is to identify possible patterns and relationships in the data without making strong causal claims.

------------------------------------------------------------------------

### Question 3: Are higher vaccination rates associated with lower numbers of ICU patients in Germany?

**Claim type:** Exploratory

We explore whether increased vaccination coverage is related to lower numbers of COVID-19 patients in intensive care units. This analysis may provide insights into the relationship between vaccination uptake and healthcare system burden.

## Audience and Context

This analysis is conducted in an academic setting as part of a data science and statistical programming project. The goal is to explore COVID-19 vaccination trends in Germany and investigate possible relationships between vaccination coverage, mortality, and intensive care unit (ICU) occupancy.

The primary audience consists of students, instructors, and individuals interested in understanding the development of vaccination campaigns during the COVID-19 pandemic. The analysis aims to communicate findings in a clear, accessible, and data-driven manner.

The results will be presented through a GitHub-hosted website created with Quarto. Appropriate presentation formats include interactive and static visualizations, time-series plots, scatter plots, statistical summaries, and brief written interpretations. This format allows the audience to explore the results and understand the key findings through both text and graphics.

## Domain-informed quality checks

We examined the key variables used in our analysis.

-   Vaccination rates (`people_vaccinated_per_hundred` and `people_fully_vaccinated_per_hundred`) range from 0 to approximately 78 percent, which is within plausible bounds.
-   ICU patient counts range from 100 to 5761 and contain no negative values.
-   Daily death counts are non-negative, although some unusually large values appear and may reflect reporting corrections or delayed reporting.
-   Several variables contain missing values. Vaccination variables have more than 800 missing observations, while ICU patient counts and death counts also contain missing data. These missing values are likely related to changes in reporting availability over time. Some variables are completely missing for Germany and will therefore not be used in the analysis.

##EDA-Plots
```{r}
library(ggplot2)

ggplot(covid_germany, aes(x = new_deaths_smoothed)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Daily COVID-19 deaths",
    x = "Daily deaths (7-day average)",
    y = "Count"
  )

```
The distribution of daily COVID-19 deaths is right-skewed. Most observations correspond to relatively low numbers of deaths, while a small number of days experienced substantially higher mortality. This indicates considerable variation in COVID-19 mortality over time.

```{r}
ggplot(covid_germany, aes(x = date)) +
  geom_line(aes(y = new_cases_smoothed, colour = "Cases")) +
  geom_line(aes(y = new_deaths_smoothed * 100, colour = "Deaths x 100")) +
  scale_colour_manual(
    values = c("Cases" = "blue", "Deaths x 100" = "red")
  ) +
  labs(
    title = "COVID-19 cases and deaths over time",
    x = "Date",
    y = "Daily cases / deaths x 100",
    colour = ""
  )
```
Both COVID-19 cases and deaths exhibit clear wave-like patterns over time. Peaks in deaths generally occur shortly after peaks in reported cases, suggesting a temporal relationship between infection levels and mortality.
