# ***Proposal for StatProg2 Projekt***

#### by Lukas Seibold and Felix Obermaier

## Dataset and research questions:

-   **Source and licence:**

    -   The Data set is from the following website: [OWID - Covid Dataset](https://ourworldindata.org/explorers/covid?Metric=Excess+mortality+%28estimates%29&Interval=Cumulative&Relative+to+population=true&country=USA~BRA~JPN~DEU)

    -   It can be downloaded in the following github repository: [OWID - github repository COVID](https://github.com/owid/covid-19-data/tree/master/public/data)

    -   The dataset is called owid-covid-data.csv

    -   The Dataset contains all Data on Covid in the years 2020 - 2024 in all countries of the world

    -   We decided we only want to look at german Data on covid so we filtered it by country

```{r}
library(readr)
library(tidyverse)

covid_germany = read_delim("data/covid_germany.csv")

```

-   **Licence:**

    -   The *Our World in Data Covid Dataset* combines COVID-19 data from several official and scientific sources, including:

        -   [Johns Hopkins University (JHU CSSE)](https://github.com/CSSEGISandData/COVID-19?utm_source=chatgpt.com) for confirmed cases and deaths

        -   official national health authorities for vaccination and hospitalization data

        -   the [Oxford COVID‑19 Government Response Tracker](https://www.bsg.ox.ac.uk/research/research-projects/covid-19-government-response-tracker?utm_source=chatgpt.com) for policy measures

        -   international organizations such as the UN, World Bank, OECD, and IHME for demographic and socioeconomic indicators.

        The dataset is published under the **Creative Commons Attribution (CC BY) license**.

        This license allows users to:

        -   use the data freely,

        -   copy and redistribute it,

        -   modify and adapt it,

        -   use it commercially,

        -   and include it in research or presentations,

        as long as proper attribution to Our World in Data and the original data sources is provided.

## What is in the data set?

The data set contains daily COVID-19 observations for Germany from January 2020 until August 2024.

-   Number of rows: 1674
-   Number of columns: 67

Each row represents one day of reported COVID-19 data in Germany.

The data set includes information on:

-   COVID-19 cases (e.g. `total_cases`, `new_cases`, `new_cases_smoothed`)
-   COVID-19 deaths (e.g. `total_deaths`, `new_deaths`)
-   Reproduction rate (`reproduction_rate`)
-   Hospital and ICU patients (`hosp_patients`, `icu_patients`)
-   Vaccination indicators
-   Testing statistics
-   Socioeconomic indicators and demographic variables

The key variables we expect to use are:

-   `date`
-   `new_cases`
-   `new_cases_smoothed`
-   `total_cases`
-   `new_deaths`
-   `total_deaths`
-   `reproduction_rate`

## Research Questions

### Question 1: How did vaccination coverage in Germany develop during the COVID-19 pandemic?

**Claim type:** Descriptive

We aim to describe how vaccination coverage changed between 2020 and 2024. In particular, we want to identify important phases of the vaccination campaign and examine how quickly vaccination rates increased over time.

To address this question, we will analyse the following variables:

-   `date`
-   `new_vaccinations(_smoothed)`,
-   `people_fully_vaccinated(_smoothed)_per_hundred`

and similar ones.

------------------------------------------------------------------------

### Question 2: How did vaccination coverage relate to different dimensions of COVID-19 severity in Germany?

**Claim type:** Exploratory

We investigate whether periods with higher COVID-19 vaccination coverage were associated with lower levels of severe COVID-19 outcomes in Germany. Rather than focusing on a single indicator, we examine multiple dimensions of disease severity, including mortality and intensive care unit (ICU) occupancy. The goal is to explore patterns and relationships in the data without making causal claims about the effects of vaccination.

To address this question, we will analyse the following variables:

-   `date`

-   `people_vaccinated_per_hundred`

-   `new_deaths_smoothed`

-   `icu_patients`

-   `new_cases_smoothed` (to provide additional context on infection dynamics)

### Sub-question 2a

Are higher vaccination rates associated with lower COVID-19 mortality?

### Sub-question 2b

Are higher vaccination rates associated with lower ICU occupancy?

## Audience and Context

This analysis is conducted in an academic setting as part of a data science and statistical programming project. The goal is to explore COVID-19 vaccination trends in Germany and investigate possible relationships between vaccination coverage, mortality, and intensive care unit (ICU) occupancy.

The primary audience consists of students, instructors, and individuals interested in understanding the development of vaccination campaigns during the COVID-19 pandemic. The analysis aims to communicate findings in a clear, accessible, and data-driven manner.

The results will be presented through a GitHub-hosted website created with Quarto. Appropriate presentation formats include interactive and static visualizations, time-series plots, scatter plots, statistical summaries, and brief written interpretations. This format allows the audience to explore the results and understand the key findings through both text and graphics.

## Domain-informed quality checks

We examined the key variables used in our analysis.

-   Vaccination rates (`people_vaccinated_per_hundred` and `people_fully_vaccinated_per_hundred`) range from 0 to approximately 78 percent, which is within plausible bounds.
-   ICU patient counts range from 100 to 5761 and contain no negative values.
-   Daily death counts are non-negative, although some unusually large values appear and may reflect reporting corrections or delayed reporting.
-   Several variables contain missing values. Vaccination variables have more than 800 missing observations, while ICU patient counts and death counts also contain missing data. These missing values are likely related to changes in reporting availability over time. Some variables are completely missing for Germany and will therefore not be used in the analysis.

## Initial Data Analysis (IDA)

#### Structure of the data

We use `glimpse()` to confirm the dimensions and data types of our key variables.

```{r}
#| eval: false
covid_germany %>%
  select(
    date, total_cases, new_cases, new_cases_smoothed,
    total_deaths, new_deaths, new_deaths_smoothed,
    reproduction_rate, icu_patients, hosp_patients,
    people_vaccinated_per_hundred, people_fully_vaccinated_per_hundred,
    stringency_index
  ) %>%
  glimpse()
```

The structure matches our expectations: `date` is read as a proper `date` type and all measurement and count variables are numeric. Cumulative variables (`total_cases`, `total_deaths`) are present from the start, while smoothed, reproduction, ICU, and vaccination variables only begin later in the pandemic, which is reflected in the leading `NA` values.

#### Summary statistics for key variables

```{r}
#| eval: false
covid_germany %>%
  select(
    new_cases, new_cases_smoothed, new_deaths, new_deaths_smoothed,
    reproduction_rate, icu_patients,
    people_vaccinated_per_hundred, people_fully_vaccinated_per_hundred,
    stringency_index
  ) %>%
  summary()
```

The table below reports the main summary statistics together with the number of missing values per variable.

| Variable                              | Min  | Median | Mean   | Max       | NA  |
|---------------------------------------|------|--------|--------|-----------|-----|
| `new_cases`                           | 0    | 0      | 29,983 | 1,588,891 | 392 |
| `new_cases_smoothed`                  | 0    | 11,167 | 30,100 | 226,984   | 397 |
| `new_deaths`                          | 0    | 0      | 136.5  | 6,460     | 392 |
| `new_deaths_smoothed`                 | 0    | 79.0   | 137.0  | 922.9     | 397 |
| `reproduction_rate`                   | 0.52 | 1.07   | 1.10   | 3.00      | 637 |
| `icu_patients`                        | 100  | 1,253  | 1,771  | 5,761     | 480 |
| `people_vaccinated_per_hundred`       | 0.03 | 77.1   | 62.5   | 77.8      | 842 |
| `people_fully_vaccinated_per_hundred` | 0.00 | 75.1   | 58.9   | 76.2      | 861 |
| `stringency_index`                    | 0    | 43.1   | 44.2   | 85.2      | 582 |

#### Plots

-   

```{r}
library(ggplot2)  
ggplot(covid_germany, aes(x = new_deaths_smoothed)) +   
  geom_histogram(bins = 30) +   
  labs(     title = "Distribution of Daily COVID-19 deaths",     x = "Daily deaths (7-day average)",     y = "Count"   ) 
```

The distribution of daily COVID-19 deaths is right-skewed. Most observations correspond to relatively low numbers of deaths, while a small number of days experienced substantially higher mortality. This indicates considerable variation in COVID-19 mortality over time.

```{r}
ggplot(covid_germany, aes(x = date)) +   
  geom_line(aes(y = new_cases_smoothed, colour = "Cases")) +   
  geom_line(aes(y = new_deaths_smoothed * 100, colour = "Deaths x 100")) +   
  scale_colour_manual(     values = c("Cases" = "blue", "Deaths x 100" = "red")   ) +   
  labs(     title = "COVID-19 cases and deaths over time",     x = "Date",     y = "Daily cases / deaths x 100",     colour = ""   )
```

Both COVID-19 cases and deaths exhibit clear wave-like patterns over time. Peaks in deaths generally occur shortly after peaks in reported cases, suggesting a temporal relationship between infection levels and mortality.

#### Data quality issues and planned handling

-   **Substantial missingness in key variables.** Vaccination variables have over 840 missing values, ICU counts 480, and the reproduction rate 637. These gaps are concentrated at the beginning of the period, before the corresponding indicators were collected. We will restrict each analysis to the time window in which the relevant variable is actually available rather than imputing early values.

-   **`hosp_patients` is entirely missing for Germany (1,674 NAs).** Although listed as a key variable in our earlier draft, it cannot be used. We will drop it and rely on `icu_patients` as the measure of hospital burden.

-   **Reporting spikes in raw counts.** `new_cases` and `new_deaths` contain extreme single-day values (up to 1.59 M cases and 6,460 deaths) caused by delayed or batched reporting. We will use the smoothed 7-day-average variables throughout to reduce the influence of these artefacts.

-   **Suspicious ICU minimum.** The minimum ICU value is exactly 100, which is more likely a lower bound or rounding artefact in the source than a true minimum. We will keep this in mind when interpreting small ICU values rather than treating it as a clean lowest observation

## Analysis plan

**Question 1 — Vaccination coverage over time (descriptive).** Variables: `date`, `people_vaccinated_per_hundred`, `people_fully_vaccinated_per_hundred`. We will use a time-series line plot to identify the phases of the vaccination campaign (start in early 2021, steep rise, plateau near 78 %). Cleaning: filter to the period where vaccination data exist and drop leading `NA`s. Main risk: vaccination figures are not reported on every single day, so minor jumps may be reporting artefacts rather than real changes.

**Question 2 — Vaccination vs. mortality (exploratory).** Variables: `people_fully_vaccinated_per_hundred` against `new_deaths_smoothed`. We will use a scatterplot (with time encoded by colour) plus a smoothed trend line (`geom_smooth`). Cleaning: restrict to the vaccination period and use the smoothed death series to dampen the reporting spikes identified above. Main risk: strong confounding by time — higher vaccination coverage coincides with new variants, seasonality, and relaxed restrictions (`stringency_index`). We therefore make no causal claims.

**Question 3 — Vaccination vs. ICU occupancy (exploratory).** Variables: `people_fully_vaccinated_per_hundred` against `icu_patients`, with the same approach and cleaning as Question 2. Same time-confounding risk applies, plus the ICU lower-bound issue noted in the quality checks, which we account for when interpreting low values.

## Workflow organisation

-   **Code style:** We follow the tidyverse style guide and enforce it by formatting all code with `styler` before committing.

-   **Packages:** `tidyverse` (including `readr`, `dplyr`, `ggplot2`)

-   **Git workflow:** We collaborate via GitHub. Each of us works on their own feature branch, and changes are merged into `main` through pull requests with a short review. Where possible we split work by research question to minimise merge conflicts.

-   **`.gitignore`:** We exclude large raw data files, rendered Quarto output, and any `.env`/secret files to keep the repository lightweight and clean.
