---
title: "test covid data"
format: html
editor: visual
---

## Test

```{r}
library(tidyverse)
library(readr)

covid_germany = read_delim("covid_germany.csv")
view(covid_germany)
```

```{r}
install.packages("skimr")
library(skimr)

skim = skim(covid_germany)
view(skim)
```

```{r}
install.packages("naniar")
library(naniar)

# Summary of missings by variable
miss_var_summary(covid_germany)

# Visualise missing pattern
vis_miss(covid_germany)
```

Für Aufgabe 3:

Einige Variablen haben **gar keine Daten** für Deutschland (`hosp_patients`, `extreme_poverty`, `handwashing_facilities` → komplett schwarz).

- Impfvariablen haben erst später Werte → unten grau, oben schwarz.

- ICU-Variablen haben viele, aber nicht alle Missing Values.

- Die Missing Values treten häufig **blockweise über die Zeit** auf und nicht zufällig.

```{r}
summary(covid_germany$people_vaccinated_per_hundred)

summary(covid_germany$people_fully_vaccinated_per_hundred)

summary(covid_germany$icu_patients)

summary(covid_germany$new_deaths)
```

```{r}
install.packages("patchwork")
```

```{r}
library(dplyr)
library(tidyverse)
```

## Ausarbeitung der Research Questions:

### RQ1:

**Model specification — RQ1**

- **Response:** `people_fully_vaccinated_per_hundred`

- **Predictor:** `date`

- **Filter:** restricted to rows where `people_fully_vaccinated_per_hundred` is not missing (vaccination reporting only begins in early 2021)

- **Method:** `geom_line` (raw trajectory) plus a LOESS-smoothed trend line (`geom_smooth`) as a trial, to assess whether smoothing adds clarity over the raw reporting data

- **Expected output:** a time-series plot showing the trajectory of full vaccination coverage, from which we identify and qualitatively describe the main campaign phases (onset, steep rise, plateau near 78%)

- **Interpretation:** the plot characterizes how full vaccination coverage evolved over time; no comparison to other variables or causal claims are made at this stage

Can the Data actually support this? - **Sanity Check:**

```{r}
glimpse(covid_germany)
```

We confirmed that `date`, `people_fully_vaccinated_per_hundred` and `new_vaccinations_smoothed`have the correct date types in order for us to create the planned line plot (Date and double) using glimpse().

```{r}
#check variable people_fully_vaccinated_per_hundred
covid_germany %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  summarise(start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  arrange(date) %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  summarise(max_gap = max(gap, na.rm = TRUE))
```

The Data coverage on vaccination stars in January of 2021 and ends in July of 2023. In this time range the data coverage is very good with a maximum gap of 1 day without coverage.

```{r}
#check variable new_vaccinations_smoothed
covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  summarise(start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  arrange(date) %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  summarise(max_gap = max(gap, na.rm = TRUE))
```

By using the same method as before we now found out that the data for `new_vaccinations_smoothed`ranges from December 2020 until July 2023 datewise. Also just like with `people_fully_vaccinated_per_hundred` the coverage for this variable is very good.

As we explored in the *Domain-informed quality checks* part of our Proposal, the values of `people_fully_vaccinated_per_hundred` ranges from 0 to 78, which is enough variation to detect distinct phases over the available period.

```{r}
covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  summarise(
    min = min(new_vaccinations_smoothed),
    max = max(new_vaccinations_smoothed),
    mean = mean(new_vaccinations_smoothed),
    sd = sd(new_vaccinations_smoothed)
  )
```

The variation of values as well as the variance of `new_vaccinations_smoothed` is also going to be sufficient to answer our research question.

## covid_germany cleanen

```{r}
covid1 <- covid_germany
```

```{r}
covid2 <- covid1 %>%
  select(where(~ all(is.na(.))))
```

```{r}

```
