# ***Proposal for StatProg2 Projekt***

#### by Lukas Seibold and Felix Obermaier

## Dataset and research questions:

- **Source and licence:**

  - The Data set is from the following website: [OWID - Covid Dataset](https://ourworldindata.org/explorers/covid?Metric=Excess+mortality+%28estimates%29&Interval=Cumulative&Relative+to+population=true&country=USA~BRA~JPN~DEU)

  - It can be downloaded in the following github repository: [OWID - github repository COVID](https://github.com/owid/covid-19-data/tree/master/public/data)

  - The dataset is called owid-covid-data.csv

  - The Dataset contains all Data on Covid in the years 2020 - 2024 in all countries of the world

  - We decided we only want to look at german Data on covid so we filtered it by country

  - Then we cleaned it, by removing all variables, that only contained NAs (6 variables removed)

  - We documented the complete cleaning process in a seperate file called data_cleaning_covid.qmd, which can be found in our github repository. All of the data we used is stored there as well in the folder "data".

```{r}
library(readr)
library(tidyverse)

covid_germany <- read_delim("data/covid_germany.csv")

```

- **Licence:**

  - The *Our World in Data Covid Dataset* combines COVID-19 data from several official and scientific sources, including:

    - [Johns Hopkins University (JHU CSSE)](https://github.com/CSSEGISandData/COVID-19?utm_source=chatgpt.com) for confirmed cases and deaths

    - official national health authorities for vaccination and hospitalization data

    - the [Oxford COVID‑19 Government Response Tracker](https://www.bsg.ox.ac.uk/research/research-projects/covid-19-government-response-tracker?utm_source=chatgpt.com) for policy measures

    - international organizations such as the UN, World Bank, OECD, and IHME for demographic and socioeconomic indicators.

    The dataset is published under the **Creative Commons Attribution (CC BY) license**.

    This license allows users to:

    - use the data freely,

    - copy and redistribute it,

    - modify and adapt it,

    - use it commercially,

    - and include it in research or presentations,

    as long as proper attribution to Our World in Data and the original data sources is provided.

## What is in the data set?

The data set contains daily COVID-19 observations for Germany from January 2020 until August 2024.

- Number of rows: 1674
- Number of columns: 61

Each row represents one day of reported COVID-19 data in Germany.

The data set includes information on:

- COVID-19 cases (e.g. `total_cases`, `new_cases`, `new_cases_smoothed`)
- COVID-19 deaths (e.g. `total_deaths`, `new_deaths`)
- Reproduction rate (`reproduction_rate`)
- Hospital and ICU patients (`hosp_patients`, `icu_patients`)
- Vaccination indicators
- Testing statistics
- Socioeconomic indicators and demographic variables

The key variables we expect to use are:

- `date`
- `new_cases`
- `new_cases_smoothed`
- `total_cases`
- `new_deaths`
- `total_deaths`
- `reproduction_rate`

Throughout this project, we primarily use the smoothed variables provided in the dataset whenever they are available. Daily COVID-19 data are affected by reporting delays and systematic weekend effects, which can introduce artificial fluctuations in the time series. The smoothed variables reduce these short-term irregularities and therefore provide a better representation of the underlying trends that are relevant for our analyses. Unless otherwise stated, all analyses are based on the smoothed versions of the variables.

## Research Questions

### Question 1: How did vaccination coverage in Germany develop during the COVID-19 pandemic?

We will analyse `people_fully_vaccinated_per_hundred` together with `new_vaccinations_smoothed` and `date`, to analyse the development of COVID-19 vaccination coverage in Germany. Using line plots and descriptive statistics, we will identify the start of the vaccination campaign, periods of rapid vaccine uptake, peak vaccination activity, and the later plateau in vaccination coverage.

**Goal:** Descriptive

**Type:** Exploratory

------------------------------------------------------------------------

### Question 2: How did vaccination coverage relate to different dimensions of COVID-19 severity in Germany?

We investigate whether COVID-19 vaccination coverage was associated with different dimensions of disease severity in Germany. To explore these relationships, we focus separately on mortality and ICU occupancy without making causal claims about the effects of vaccination.

**Goal:** Exploratory

**Type:** Exploratory

### Sub-question 2a: Are higher vaccination rates associated with lower COVID-19 mortality?

We will analyse `people_vaccinated_per_hundred` together with `new_deaths_smoothed` and `date` using line plots, scatter plots, correlation analysis, and descriptive statistics to examine whether higher vaccination coverage was associated with lower COVID-19 mortality in Germany.

Since this is an exploratory analysis, we investigate associations between vaccination coverage and mortality without making strong causal claims.

### Sub-question 2b: Are higher vaccination rates associated with lower ICU occupancy?

We will analyse `people_vaccinated_per_hundred` together with `icu_patients` and `date` using line plots, scatter plots, correlation analysis, and descriptive statistics to examine whether higher vaccination coverage was associated with lower ICU occupancy in Germany.

Since this is an exploratory analysis, we investigate associations between vaccination coverage and ICU occupancy without making strong causal claims.

------------------------------------------------------------------------

## Audience and Context

This analysis is conducted in an academic setting as part of a data science and statistical programming project. The goal is to explore COVID-19 vaccination trends in Germany and investigate possible relationships between vaccination coverage, mortality, and intensive care unit (ICU) occupancy.

The primary audience consists of students, instructors, and individuals interested in understanding the development of vaccination campaigns during the COVID-19 pandemic. The analysis aims to communicate findings in a clear, accessible, and data-driven manner.

The results will be presented through a GitHub-hosted website created with Quarto. Appropriate presentation formats include interactive and static visualizations, time-series plots, scatter plots, statistical summaries, and brief written interpretations. This format allows the audience to explore the results and understand the key findings through both text and graphics.

## Domain-informed quality checks

We examined the key variables used in our analysis.

- Vaccination rates (`people_vaccinated_per_hundred` and `people_fully_vaccinated_per_hundred`) range from 0 to approximately 78 percent, which is within plausible bounds.
- ICU patient counts range from 100 to 5761 and contain no negative values.
- Daily death counts are non-negative, although some unusually large values appear and may reflect reporting corrections or delayed reporting.
- Several variables contain missing values. Vaccination variables have more than 800 missing observations, while ICU patient counts and death counts also contain missing data. These missing values are likely related to changes in reporting availability over time. Some variables are completely missing for Germany and will therefore not be used in the analysis.

## Initial Data Analysis (IDA)

#### Structure of the data

We use `glimpse()` to confirm the dimensions and data types of our key variables.

```{r}
#| eval: false
covid_germany %>%
  select(
    date, total_cases, new_cases, new_cases_smoothed,
    total_deaths, new_deaths, new_deaths_smoothed,
    reproduction_rate, icu_patients,
    people_vaccinated_per_hundred, people_fully_vaccinated_per_hundred,
    stringency_index
  ) %>%
  glimpse()
```

The structure matches our expectations: `date` is read as a proper `date` type and all measurement and count variables are numeric. Cumulative variables (`total_cases`, `total_deaths`) are present from the start, while smoothed, reproduction, ICU, and vaccination variables only begin later in the pandemic, which is reflected in the leading `NA` values.

#### Summary statistics for key variables

```{r}
#| eval: false
covid_germany %>%
  select(
    new_cases, new_cases_smoothed, new_deaths, new_deaths_smoothed,
    reproduction_rate, icu_patients,
    people_vaccinated_per_hundred, people_fully_vaccinated_per_hundred,
    stringency_index
  ) %>%
  summary()
```

The table below reports the main summary statistics together with the number of missing values per variable.

| Variable                              | Min  | Median | Mean   | Max       | NA  |
|---------------------------------------|------|--------|--------|-----------|-----|
| `new_cases`                           | 0    | 0      | 29,983 | 1,588,891 | 392 |
| `new_cases_smoothed`                  | 0    | 11,167 | 30,100 | 226,984   | 397 |
| `new_deaths`                          | 0    | 0      | 136.5  | 6,460     | 392 |
| `new_deaths_smoothed`                 | 0    | 79.0   | 137.0  | 922.9     | 397 |
| `reproduction_rate`                   | 0.52 | 1.07   | 1.10   | 3.00      | 637 |
| `icu_patients`                        | 100  | 1,253  | 1,771  | 5,761     | 480 |
| `people_vaccinated_per_hundred`       | 0.03 | 77.1   | 62.5   | 77.8      | 842 |
| `people_fully_vaccinated_per_hundred` | 0.00 | 75.1   | 58.9   | 76.2      | 861 |
| `stringency_index`                    | 0    | 43.1   | 44.2   | 85.2      | 582 |

#### Plots

```{r}
#| echo: false
library(ggplot2)  
ggplot(covid_germany, aes(x = new_deaths_smoothed)) +   
  geom_histogram(bins = 30) +   
  labs(     title = "Distribution of Daily COVID-19 deaths",     x = "Daily deaths (7-day average)",     y = "Count"   ) 
```

The distribution of daily COVID-19 deaths is right-skewed. Most observations correspond to relatively low numbers of deaths, while a small number of days experienced substantially higher mortality. This indicates considerable variation in COVID-19 mortality over time.

```{r}
#| echo: false
ggplot(covid_germany, aes(x = date)) +   
  geom_line(aes(y = new_cases_smoothed, colour = "Cases")) +   
  geom_line(aes(y = new_deaths_smoothed * 100, colour = "Deaths x 100")) +   
  scale_colour_manual(     values = c("Cases" = "blue", "Deaths x 100" = "red")   ) +   
  labs(     title = "COVID-19 cases and deaths over time",     x = "Date",     y = "Daily cases / deaths x 100",     colour = ""   )
```

Both COVID-19 cases and deaths exhibit clear wave-like patterns over time. Peaks in deaths generally occur shortly after peaks in reported cases, suggesting a temporal relationship between infection levels and mortality.

**Note:** The number of daily deaths has been multiplied by a factor of 100 to make both time series visible on the same scale. This transformation is only for visualization purposes.

#### Data quality issues and planned handling

- **Substantial missingness in key variables.** Vaccination variables have over 840 missing values, ICU counts 480, and the reproduction rate 637. These gaps are concentrated at the beginning of the period, before the corresponding indicators were collected. We will restrict each analysis to the time window in which the relevant variable is actually available rather than imputing early values.

- **`hosp_patients` is entirely missing for Germany (1,674 NAs).** Although listed as a key variable in our earlier draft, it cannot be used. We will drop it and rely on `icu_patients` as the measure of hospital burden.

- **Reporting spikes in raw counts.** `new_cases` and `new_deaths` contain extreme single-day values (up to 1.59 M cases and 6,460 deaths) caused by delayed or batched reporting. We will use the smoothed 7-day-average variables throughout to reduce the influence of these artefacts.

- **Suspicious ICU minimum.** The minimum ICU value is exactly 100, which is more likely a lower bound or rounding artefact in the source than a true minimum. We will keep this in mind when interpreting small ICU values rather than treating it as a clean lowest observation

## **Analysis plan**

**RQ1 — Vaccination coverage over time (descriptive).** Variables: `date`, `people_fully_vaccinated_per_hundred`, `new_vaccinations_smoothed`. We will use a time-series line plot of `people_fully_vaccinated_per_hundred` to identify the overall phases of the vaccination campaign (start in early 2021, steep rise, plateau near 78 %), together with a line plot of `new_vaccinations_smoothed` to identify the period of peak vaccination activity, since the cumulative coverage variable is monotonic and cannot show a peak by itself. We will supplement both plots with descriptive statistics (e.g. range, mean, and timing of maximum daily vaccinations). Cleaning: filter each variable separately to the period where it is non-missing, since `people_fully_vaccinated_per_hundred` and `new_vaccinations_smoothed` may have different missingness patterns. Main risk: vaccination figures are not reported on every single day, so minor jumps or short gaps may be reporting artefacts rather than real changes in uptake.

**RQ2a — Vaccination vs. mortality (exploratory).** Variables: `people_vaccinated_per_hundred`, `new_deaths_smoothed`, `date`. We will use line plots, scatterplots, correlation analysis, and descriptive statistics to examine whether higher vaccination coverage was associated with lower COVID-19 mortality in Germany. Cleaning: restrict to the period where vaccination data exist and use the smoothed death series to dampen reporting spikes identified earlier. Main risk: strong confounding by time — higher vaccination coverage coincides with new variants, seasonality, and relaxed restrictions (`stringency_index`). Since this is an exploratory analysis, we examine associations only and make no causal claims.

**RQ2b — Vaccination vs. ICU occupancy (exploratory).** Variables: `people_vaccinated_per_hundred`, `icu_patients`, `date`. We will use the same approach as RQ2a (line plots, scatterplots, correlation analysis, descriptive statistics) to examine whether higher vaccination coverage was associated with lower ICU occupancy in Germany. Cleaning: same as RQ2a. Same time-confounding risk applies, plus the ICU lower-bound artefact (minimum value of exactly 100) noted in the quality checks, which we account for when interpreting low values. As an exploratory analysis, we make no causal claims.

## Workflow organisation

- **Code style:** We follow the tidyverse style guide and enforce it by formatting all code with `styler` before committing.

- **Packages:** `tidyverse` (including `readr`, `dplyr`, `ggplot2`)

- **Git workflow:** We collaborate via GitHub. Each of us works on their own feature branch, and changes are merged into `main` through pull requests with a short review. Where possible we split work by research question to minimise merge conflicts.

- **`.gitignore`:** We exclude large raw data files, rendered Quarto output, and any `.env`/secret files to keep the repository lightweight and clean.

## Answering the Research Questions

### Research Question 1

#### **Model specification — RQ1**

- **Response:** `people_fully_vaccinated_per_hundred`

- **Predictor:** `date`

- **Filter:** restricted to rows where `people_fully_vaccinated_per_hundred` is not missing (vaccination reporting only begins in early 2021)

- **Method:** `geom_line` (raw trajectory) plus a LOESS-smoothed trend line (`geom_smooth`) as a trial, to assess whether smoothing adds clarity over the raw reporting data

- **Expected output:** a time-series plot showing the trajectory of full vaccination coverage, from which we identify and qualitatively describe the main campaign phases (onset, steep rise, plateau near 78%)

- **Interpretation:** the plot characterizes how full vaccination coverage evolved over time; no comparison to other variables or causal claims are made at this stage

#### Can the data support our RQ1? – Sanity Check

```{r}
glimpse(covid_germany)
```

We confirmed that `date`, `people_fully_vaccinated_per_hundred` and `new_vaccinations_smoothed`have the correct data types in order for us to create the planned line plot (Date and double) using glimpse().

```{r}
#check variable people_fully_vaccinated_per_hundred
covid_germany %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  summarise(start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  arrange(date) %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  summarise(max_gap = max(gap, na.rm = TRUE))
```

The Data coverage on vaccination starts in January of 2021 and ends in July of 2023. In this time range the data coverage is very good with a maximum gap of 1 day without coverage.

```{r}
#check variable new_vaccinations_smoothed
covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  summarise(start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  arrange(date) %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  summarise(max_gap = max(gap, na.rm = TRUE))
```

By using the same method as before we now found out that the data for `new_vaccinations_smoothed`ranges from December 2020 until July 2023 datewise. Also just like with `people_fully_vaccinated_per_hundred` the coverage for this variable is very good.

As we explored in the *Domain-informed quality checks* part of our Proposal, the values of `people_fully_vaccinated_per_hundred` ranges from 0 to 78, which is enough variation to detect distinct phases over the available period.

```{r}
covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  summarise(
    min = min(new_vaccinations_smoothed),
    max = max(new_vaccinations_smoothed),
    mean = mean(new_vaccinations_smoothed),
    sd = sd(new_vaccinations_smoothed)
  )
```

The variation of values as well as the variance of `new_vaccinations_smoothed` is also going to be sufficient to answer our research question.

### Research Question 2

#### **Model specification — RQ2a**

- **Response:** `new_deaths_smoothed` — daily new confirmed COVID-19 deaths, 7-day smoothed (`<dbl>`)

- **Predictor:** `people_vaccinated_per_hundred` — share of people with at least one dose (`<dbl>`)

- **Filter:** restricted to rows where both variables are not missing (vaccination reporting only begins in late December 2020, which is the binding constraint)

- **Method:** Pearson correlation between the two variables, tested with the theory-based `{infer}` pipeline (`specify() → assume("t")` / `calculate(stat = "correlation")`), plus a scatter plot to inspect the shape of the relationship

- **Expected output:** the observed correlation coefficient, a *p*-value, a 95% confidence interval, and a scatter plot of mortality against vaccination coverage

- **Interpretation:** the sign of the coefficient tells us whether days with higher vaccination coverage tended to have lower mortality; because both variables move over time, the association is exploratory and no causal claims are made

#### **Model specification — RQ2b**

- **Response:** `icu_patients` — number of COVID-19 patients in intensive care on a given day (`<dbl>`)

- **Predictor:** `people_vaccinated_per_hundred` — share of people with at least one dose (`<dbl>`)

- **Filter:** restricted to rows where both variables are not missing (same binding constraint as RQ2a)

- **Method:** same as RQ2a — Pearson correlation via the theory-based `{infer}` pipeline, plus a scatter plot

- **Expected output:** the observed correlation coefficient, a *p*-value, a 95% confidence interval, and a scatter plot of ICU occupancy against vaccination coverage

- **Interpretation:** the sign of the coefficient tells us whether days with higher vaccination coverage tended to have lower ICU occupancy; the ICU lower-bound artefact (minimum value of exactly 100, see quality checks) is kept in mind when interpreting low values, and no causal claims are made

#### Can the data support our RQ2? – Sanity Check

As for RQ1, we first confirm that the variables exist with the correct types using `glimpse()` (see the RQ1 sanity check above): `people_vaccinated_per_hundred`, `new_deaths_smoothed` and `icu_patients` are all doubles, and `date` is a Date, which is what the planned correlation analysis and scatter plots require.

Unlike RQ1, RQ2 compares **two variables at a time**, so the decisive question is not only how well each variable is covered, but whether their reporting periods **overlap**. We therefore check the coverage window of each variable and then the size of the overlap.

```{r}
covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred)) %>%
  summarise(n = n(), start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred)) %>%
  arrange(date) %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  summarise(max_gap = max(gap, na.rm = TRUE))
```

The vaccination coverage data ranges from late December 2020 until April 2023 (832 days) with a maximum gap of 1 day, so the coverage within this window is very good.

```{r}
covid_germany %>%
  filter(!is.na(new_deaths_smoothed)) %>%
  summarise(n = n(), start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(icu_patients)) %>%
  summarise(n = n(), start = min(date), end = max(date))
```

Both response variables cover a longer period than the vaccination data: `new_deaths_smoothed` runs from January 2020 to July 2023 and `icu_patients` from March 2020 to June 2023, each without gaps larger than 1 day. The vaccination variable is therefore the binding constraint for both sub-questions.

```{r}
covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred),
         !is.na(new_deaths_smoothed)) %>%
  summarise(n = n(), start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred),
         !is.na(icu_patients)) %>%
  summarise(n = n(), start = min(date), end = max(date))
```

The overlap is identical for both sub-questions: all 832 days with vaccination data also have mortality and ICU data. This means RQ2a and RQ2b can be analysed on the same time window (2020-12-27 to 2023-04-07), which keeps the two analyses directly comparable.

```{r}
covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred),
         !is.na(new_deaths_smoothed),
         !is.na(icu_patients)) %>%
  summarise(
    vacc_min = min(people_vaccinated_per_hundred),
    vacc_max = max(people_vaccinated_per_hundred),
    deaths_min = min(new_deaths_smoothed),
    deaths_max = max(new_deaths_smoothed),
    icu_min = min(icu_patients),
    icu_max = max(icu_patients)
  )
```

Within the window, vaccination coverage rises from close to 0 to about 78 per hundred, smoothed daily deaths range from about 4 to over 800, and ICU occupancy from about 350 to over 5700 patients. All three variables show more than enough variation to estimate a correlation, so the data can support both sub-questions as specified.

## Analysis

### Research Question 1

```{r}
#| fig-alt: "Line chart showing the share of people fully vaccinated against COVID-19 in Germany from late 2020 to 2023. The line starts near zero in early 2021, rises steeply through 2021, and then levels off at around 76 percent."
covid_germany %>% filter(!is.na(people_fully_vaccinated_per_hundred)) %>% ggplot(aes(
  x = date,
  y = people_fully_vaccinated_per_hundred
))+
  geom_line()+
  labs(title = "Development of vaccination coverage in Germany during Covid",
       x = NULL,
       y = "People fully vaccinated per hundred",
       caption = "Source: Our World in Data COVID-19 dataset")+
  theme_minimal()
  
```

### Research Question 2

#### Inference — correlation test with `{infer}`

We prepare one analysis dataset for both sub-questions, restricted to the overlap window from the sanity check:

```{r}
rq2_data <- covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred),
         !is.na(new_deaths_smoothed),
         !is.na(icu_patients)) %>%
  select(date, people_vaccinated_per_hundred,
         new_deaths_smoothed, icu_patients)
```

**RQ2a: vaccination coverage and mortality**

```{r}
# observed correlation
obs_cor_deaths <- rq2_data %>%
  specify(new_deaths_smoothed ~ people_vaccinated_per_hundred) %>%
  calculate(stat = "correlation")

obs_cor_deaths

# theoretical null distribution and p-value
null_dist_deaths <- rq2_data %>%
  specify(new_deaths_smoothed ~ people_vaccinated_per_hundred) %>%
  assume("t")

null_dist_deaths %>%
  get_p_value(obs_stat = obs_cor_deaths, direction = "two-sided")

# confidence interval from the equivalent base R test
cor.test(rq2_data$new_deaths_smoothed,
         rq2_data$people_vaccinated_per_hundred)
```

**RQ2b: vaccination coverage and ICU occupancy**

```{r}
obs_cor_icu <- rq2_data %>%
  specify(icu_patients ~ people_vaccinated_per_hundred) %>%
  calculate(stat = "correlation")

obs_cor_icu

null_dist_icu <- rq2_data %>%
  specify(icu_patients ~ people_vaccinated_per_hundred) %>%
  assume("t")

null_dist_icu %>%
  get_p_value(obs_stat = obs_cor_icu, direction = "two-sided")

cor.test(rq2_data$icu_patients,
         rq2_data$people_vaccinated_per_hundred)
```

For RQ2a we observe a moderate negative correlation of about r = -0.53 (95% CI roughly -0.58 to -0.48), and for RQ2b a somewhat stronger negative correlation of about r = -0.65 (95% CI roughly -0.69 to -0.61). Both *p*-values are far below any conventional threshold, so a correlation of zero is not compatible with the data.

**One-sentence interpretation (Ex 2.4):** In the period where vaccination data exist, days with higher vaccination coverage tended to have lower COVID-19 mortality and lower ICU occupancy — an exploratory, associational finding that is consistent with our research questions but does not establish a causal effect of vaccination.

## Results

### Research Question 1

### Research Question 2

#### Answering RQ2a: Are higher vaccination rates associated with lower COVID-19 mortality?

```{r} #| fig-alt: "Scatter plot of smoothed daily COVID-19 deaths in Germany against the share of people vaccinated per hundred, from late 2020 to early 2023. Points are coloured by date. Deaths are high at low vaccination coverage in early 2021, drop in mid 2021, rise again during the winter waves, and are low at high coverage in 2022 and 2023. The overall trend is downward." rq2_data %>%   ggplot(aes(x = people_vaccinated_per_hundred,              y = new_deaths_smoothed,              colour = date)) +   geom_point(alpha = 0.6) +   labs(title = "Vaccination coverage and COVID-19 mortality in Germany",        subtitle = "Each point is one day; colour shows the date",        x = "People vaccinated per hundred",        y = "New deaths (7-day smoothed)",        colour = NULL,        caption = "Source: Our World in Data COVID-19 dataset") +   theme_minimal()}
```

The scatter plot shows the negative association quantified above (r ≈ -0.53): days with higher vaccination coverage tended to have lower smoothed daily deaths. The date colouring also makes the limitation of this comparison visible. Because vaccination coverage only ever increases, the x-axis is closely tied to calendar time: points on the left are from early 2021 and points on the right from 2022/23. The high-mortality points at low coverage belong to the winter wave of early 2021, while the winter wave of 2021/22 appears as a band of elevated deaths at around 60–70 per hundred. So the negative correlation mixes two things we cannot separate here: the protective association of vaccination and everything else that changed over time (variants, seasonality, and restrictions captured in `stringency_index`).

**Answer to RQ2a:** Yes — in our data, higher vaccination rates are associated with lower COVID-19 mortality (r ≈ -0.53, 95% CI -0.58 to -0.48). As stated in our analysis plan, this is an exploratory, associational result: because vaccination coverage rises monotonically over time, calendar time is a strong confounder, and we do not interpret this association causally.

#### Answering RQ2b: Are higher vaccination rates associated with lower ICU occupancy?

```{r} #| fig-alt: "Scatter plot of the number of COVID-19 ICU patients in Germany against the share of people vaccinated per hundred, from late 2020 to early 2023. Points are coloured by date. ICU occupancy is around 3000 to 5700 patients at low vaccination coverage in early 2021, falls in mid 2021, peaks again near 5000 at around 65 per hundred during the winter wave of 2021, and is mostly below 2000 at high coverage in 2022 and 2023. The overall trend is downward." rq2_data %>%   ggplot(aes(x = people_vaccinated_per_hundred,              y = icu_patients,              colour = date)) +   geom_point(alpha = 0.6) +   labs(title = "Vaccination coverage and ICU occupancy in Germany",        subtitle = "Each point is one day; colour shows the date",        x = "People vaccinated per hundred",        y = "COVID-19 patients in intensive care",        colour = NULL,        caption = "Source: Our World in Data COVID-19 dataset") +   theme_minimal()}
```

The association is negative and somewhat stronger than for mortality (r ≈ -0.65): days with higher vaccination coverage tended to have fewer COVID-19 patients in intensive care. The same time-confounding caveat applies as in RQ2a. In addition, our domain-informed quality checks flagged that the minimum ICU value of exactly 100 looks like a reporting lower bound; since the values in the overlap window start at around 350 patients, this artefact does not affect the analysis here, but we note it for completeness.

**Answer to RQ2b:** Yes — in our data, higher vaccination rates are associated with lower ICU occupancy (r ≈ -0.65, 95% CI -0.69 to -0.61). As with RQ2a, this is an exploratory, associational finding and not evidence of a causal effect, because vaccination coverage and calendar time are inseparable in this dataset.

#### How stable is the association over time?

To make the time-confounding limitation concrete rather than only stating it, we re-computed the correlations within calendar years:

```{r} rq2_data %>%   mutate(year = factor(format(date, "%Y"))) %>%   group_by(year) %>%   summarise(     n = n(),     cor_deaths = cor(people_vaccinated_per_hundred, new_deaths_smoothed),     cor_icu = cor(people_vaccinated_per_hundred, icu_patients)   )}
```

Within single years the correlations are noticeably weaker and unstable (for example, the ICU correlation is strong in 2022 but close to zero in early 2023, while the mortality correlation weakens in 2022). This confirms that a large part of the overall negative correlation is driven by the long-run trend — coverage rising while the pandemic waned — rather than by a stable day-to-day relationship. It is exactly the behaviour our analysis plan anticipated when it named time as the main confounding risk, and it is the reason we report RQ2 as exploratory.
