---
title: "COVID-19 Vaccination and Disease Severity in Germany"
format: html
author: Felix Obermaier, Lukas Seibold
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(infer)

covid_germany <- read_delim("data2/covid_germany.csv")
```

# Introduction

This project investigates the development of COVID-19 vaccination coverage in Germany and explores whether vaccination rates were associated with measures of disease severity during the pandemic.

We use publicly available data from the *Our World in Data* COVID-19 dataset, restricted to Germany.

# Dataset

The dataset contains daily observations for Germany between 2020 and 2024.

For our analyses we mainly use the following variables:

-   `date`
-   `people_vaccinated_per_hundred`
-   `new_deaths_smoothed`
-   `icu_patients`
-   `new_cases_smoothed` (for context)

Before the analysis, we inspected the dataset for missing values, variable types, and implausible observations.

# Research Questions

Our project addresses the following questions:

1.  How did COVID-19 vaccination coverage in Germany change over time?

2.  Was higher vaccination coverage associated with lower COVID-19 severity?

    -   Was higher vaccination coverage associated with lower COVID-19 mortality?
    -   Was higher vaccination coverage associated with lower ICU occupancy?

# Preliminary Analysis

## Vaccination Coverage - Research Question 1

#### **Model specification — RQ1**

-   **Response:** `people_fully_vaccinated_per_hundred`

-   **Predictor:** `date`

-   **Filter:** restricted to rows where `people_fully_vaccinated_per_hundred` is not missing (vaccination reporting only begins in early 2021)

-   **Method:** `geom_line` (raw trajectory)

-   **Expected output:** a time-series plot showing the trajectory of full vaccination coverage, from which we identify and qualitatively describe the main campaign phases (onset, steep rise, plateau near 78%)

-   **Interpretation:** the plot characterizes how full vaccination coverage evolved over time; no comparison to other variables or causal claims are made at this stage

#### Can the data support our RQ1? – Sanity Check

```{r}
covid_germany %>%
  select(date, people_fully_vaccinated_per_hundred, new_vaccinations_smoothed) %>%
  glimpse()
```

We confirmed that `date`, `people_fully_vaccinated_per_hundred` and `new_vaccinations_smoothed` have the correct data types in order for us to create the planned line plot (Date and double) using glimpse().

```{r}
covid_germany %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  summarise(start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  arrange(date) %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  summarise(max_gap = max(gap, na.rm = TRUE))
```

The Data coverage on vaccination starts in January of 2021 and ends in July of 2023. In this time range the data coverage is very good with a maximum gap of 1 day without coverage.

```{r}
covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  summarise(start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  arrange(date) %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  summarise(max_gap = max(gap, na.rm = TRUE))
```

By using the same method as before we now found out that the data for `new_vaccinations_smoothed` ranges from December 2020 until July 2023 datewise. Also just like with `people_fully_vaccinated_per_hundred` the coverage for this variable is very good.

As we explored in the *Domain-informed quality checks* part of our Proposal, the values of `people_fully_vaccinated_per_hundred` ranges from 0 to 78, which is enough variation to detect distinct phases over the available period.

```{r}
covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  summarise(
    min = min(new_vaccinations_smoothed),
    max = max(new_vaccinations_smoothed),
    mean = mean(new_vaccinations_smoothed),
    sd = sd(new_vaccinations_smoothed)
  )
```

The variation of values as well as the variance of `new_vaccinations_smoothed` is also going to be sufficient to answer our research question.

```{r}
covid_germany %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>% 
  ggplot(
    aes(x = date, y = people_fully_vaccinated_per_hundred)
  ) +
   geom_line(
      colour = "steelblue",
      linewidth = 1
    ) +
    labs(
      title = "COVID-19 vaccination coverage in Germany",
      subtitle = "People vaccinated per 100 inhabitants",
      x = "Date",
      y = "Vaccination coverage (%)"
    ) +
    scale_x_date(date_breaks = "2 months",
                date_labels = "%b\n%Y") +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 11),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10)
    )
```

Vaccination coverage increased rapidly after the campaign started in early 2021. During the first half of 2021 the slope is steep, indicating a fast rollout of vaccinations. Growth slowed substantially during late 2021 and eventually reached a plateau around 76–78% of the population. Afterwards only minor increases were observed until reporting ended in mid-2023.

```{r}
covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  ggplot(aes(date, new_vaccinations_smoothed)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Daily Vaccination Activity",
    subtitle = "7-day smoothed number of vaccinations",
    x = "Date",
    y = "New vaccinations"
  ) +
  theme_minimal()
```

In contrast to cumulative vaccination coverage, daily vaccination activity exhibits a pronounced peak during the first vaccination campaign in 2021. Afterwards vaccination activity declined steadily with only smaller temporary increases during later booster campaigns.

## Vaccination and COVID-19 Severity - Research Question 2

#### **Model specification — RQ2a**

-   **Response:** `new_deaths_smoothed` — daily new confirmed COVID-19 deaths, 7-day smoothed (`<dbl>`)

-   **Predictor:** `people_vaccinated_per_hundred` — share of people with at least one dose (`<dbl>`)

-   **Filter:** restricted to rows where both variables are not missing (vaccination reporting only begins in late December 2020, which is the binding constraint)

-   **Method:** Pearson correlation, tested with a permutation-based {infer} pipeline (`specify()` → `hypothesize()` → `generate()` → `calculate()`) for the p-value, complemented by `cor.test()` for the theory-based 95 % confidence interval; plus a scatter plot.

-   **Expected output:** the observed correlation coefficient, a *p*-value, a 95% confidence interval, and a scatter plot of mortality against vaccination coverage

-   **Interpretation:** the sign of the coefficient tells us whether days with higher vaccination coverage tended to have lower mortality; because both variables move over time, the association is exploratory and no causal claims are made

#### **Model specification — RQ2b**

-   **Response:** `icu_patients` — number of COVID-19 patients in intensive care on a given day (`<dbl>`)

-   **Predictor:** `people_vaccinated_per_hundred` — share of people with at least one dose (`<dbl>`)

-   **Filter:** restricted to rows where both variables are not missing (same binding constraint as RQ2a)

-   **Method:** same as RQ2a — Pearson correlation via the theory-based `{infer}` pipeline, plus a scatter plot

-   **Expected output:** the observed correlation coefficient, a *p*-value, a 95% confidence interval, and a scatter plot of ICU occupancy against vaccination coverage

-   **Interpretation:** the sign of the coefficient tells us whether days with higher vaccination coverage tended to have lower ICU occupancy; the ICU lower-bound artefact (minimum value of exactly 100, see quality checks) is kept in mind when interpreting low values, and no causal claims are made

#### Can the data support our RQ2? – Sanity Check

As for RQ1, we first confirm that the variables exist with the correct types using `glimpse()` (see the RQ1 sanity check above): `people_vaccinated_per_hundred`, `new_deaths_smoothed` and `icu_patients` are all doubles, and `date` is a Date, which is what the planned correlation analysis and scatter plots require.

Unlike RQ1, RQ2 compares **two variables at a time**, so the decisive question is not only how well each variable is covered, but whether their reporting periods **overlap**. We therefore check the coverage window of each variable and then the size of the overlap.

```{r}
covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred)) %>%
  summarise(n = n(), start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred)) %>%
  arrange(date) %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  summarise(max_gap = max(gap, na.rm = TRUE))
```

The vaccination coverage data ranges from late December 2020 until April 2023 (832 days) with a maximum gap of 1 day, so the coverage within this window is very good.

```{r}
covid_germany %>%
  filter(!is.na(new_deaths_smoothed)) %>%
  summarise(n = n(), start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(icu_patients)) %>%
  summarise(n = n(), start = min(date), end = max(date))
```

Both response variables cover a longer period than the vaccination data: `new_deaths_smoothed` runs from January 2020 to July 2023 and `icu_patients` from March 2020 to June 2023, each without gaps larger than 1 day. The vaccination variable is therefore the binding constraint for both sub-questions.

```{r}
covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred),
         !is.na(new_deaths_smoothed)) %>%
  summarise(n = n(), start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred),
         !is.na(icu_patients)) %>%
  summarise(n = n(), start = min(date), end = max(date))
```

The overlap is identical for both sub-questions: all 832 days with vaccination data also have mortality and ICU data. This means RQ2a and RQ2b can be analysed on the same time window (2020-12-27 to 2023-04-07), which keeps the two analyses directly comparable.

```{r}
covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred),
         !is.na(new_deaths_smoothed),
         !is.na(icu_patients)) %>%
  summarise(
    vacc_min = min(people_vaccinated_per_hundred),
    vacc_max = max(people_vaccinated_per_hundred),
    deaths_min = min(new_deaths_smoothed),
    deaths_max = max(new_deaths_smoothed),
    icu_min = min(icu_patients),
    icu_max = max(icu_patients)
  )
```

Within the window, vaccination coverage rises from close to 0 to about 78 per hundred, smoothed daily deaths range from about 4 to over 800, and ICU occupancy from about 350 to over 5700 patients. All three variables show more than enough variation to estimate a correlation, so the data can support both sub-questions as specified.

## Analysis

### Research Question 1

```{r}
#| fig-alt: "Line chart showing the share of people fully vaccinated against COVID-19 in Germany from late 2020 to 2023. The line starts near zero in early 2021, rises steeply through 2021, and then levels off at around 76 percent."
covid_germany %>% filter(!is.na(people_fully_vaccinated_per_hundred)) %>% ggplot(aes(
  x = date,
  y = people_fully_vaccinated_per_hundred
))+
  geom_line()+
  labs(title = "Development of vaccination coverage in Germany during Covid",
       x = NULL,
       y = "People fully vaccinated per hundred",
       caption = "Source: Our World in Data COVID-19 dataset")+
  theme_minimal()
  
```

### Research Question 2

#### Inference — correlation test with `{infer}`

We prepare one analysis dataset for both sub-questions, restricted to the overlap window from the sanity check:

```{r}
#| include: false
rq2_data <- covid_germany %>%
  filter(!is.na(people_vaccinated_per_hundred),
         !is.na(new_deaths_smoothed),
         !is.na(icu_patients)) %>%
  select(date, people_vaccinated_per_hundred,
         new_deaths_smoothed, icu_patients)
```

**RQ2a: vaccination coverage and mortality**

```{r}
# observed correlation
obs_cor_deaths <- rq2_data %>%
  specify(new_deaths_smoothed ~ people_vaccinated_per_hundred) %>%
  calculate(stat = "correlation")

obs_cor_deaths

# permutation-based null distribution
set.seed(123)

null_dist_deaths <- rq2_data %>%
  specify(new_deaths_smoothed ~ people_vaccinated_per_hundred) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "correlation")

# p-value
null_dist_deaths %>%
  get_p_value(
    obs_stat = obs_cor_deaths,
    direction = "two-sided"
  )

# theoretical Pearson correlation test and confidence interval
cor.test(
  rq2_data$new_deaths_smoothed,
  rq2_data$people_vaccinated_per_hundred
)
```

**RQ2b: vaccination coverage and ICU occupancy**

```{r}
obs_cor_icu <- rq2_data %>%
  specify(icu_patients ~ people_vaccinated_per_hundred) %>%
  calculate(stat = "correlation")

obs_cor_icu

set.seed(123)

null_dist_icu <- rq2_data %>%
  specify(icu_patients ~ people_vaccinated_per_hundred) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "correlation")

null_dist_icu %>%
  get_p_value(
    obs_stat = obs_cor_icu,
    direction = "two-sided"
  )

cor.test(
  rq2_data$icu_patients,
  rq2_data$people_vaccinated_per_hundred
)
```

For RQ2a we observe a moderate negative correlation of about r = -0.53 (95% CI roughly -0.58 to -0.48), and for RQ2b a somewhat stronger negative correlation of about r = -0.65 (95% CI roughly -0.69 to -0.61). Both *p*-values are far below any conventional threshold, so a correlation of zero is not compatible with the data.

**One-sentence interpretation (Ex 2.4):** In the period where vaccination data exist, days with higher vaccination coverage tended to have lower COVID-19 mortality and lower ICU occupancy — an exploratory, associational finding that is consistent with our research questions but does not establish a causal effect of vaccination.

