------------------------------------------------------------------------

title: "COVID-19 Vaccination and Disease Severity in Germany" format: html ------------

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(skimr)
library(infer)

covid_germany <- read_delim("data2/covid_germany.csv")
```

# Introduction

This project investigates the development of COVID-19 vaccination coverage in Germany and explores whether vaccination rates were associated with measures of disease severity during the pandemic.

We use publicly available data from the *Our World in Data* COVID-19 dataset, restricted to Germany.

# Dataset

The dataset contains daily observations for Germany between 2020 and 2024.

For our analyses we mainly use the following variables:

-   `date`
-   `people_vaccinated_per_hundred`
-   `new_deaths_smoothed`
-   `icu_patients`
-   `new_cases_smoothed` (for context)

Before the analysis, we inspected the dataset for missing values, variable types, and implausible observations.

# Research Questions

Our project addresses the following questions:

1.  How did COVID-19 vaccination coverage in Germany change over time?

2.  Was higher vaccination coverage associated with lower COVID-19 severity?

    -   Was higher vaccination coverage associated with lower COVID-19 mortality?
    -   Was higher vaccination coverage associated with lower ICU occupancy?

# Preliminary Analysis

## Vaccination Coverage

```{r}
ggplot(
  covid_germany,
  aes(x = date, y = people_vaccinated_per_hundred)
) +
  geom_line(
    colour = "#2C7FB8",
    linewidth = 1
  ) +
  labs(
    title = "COVID-19 vaccination coverage in Germany",
    subtitle = "People vaccinated per 100 inhabitants",
    x = "Date",
    y = "Vaccination coverage (%)"
  ) +
  scale_x_date(date_breaks = "6 months",
               date_labels = "%b\n%Y") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

This figure illustrates how vaccination coverage developed throughout the pandemic.

## Vaccination and Mortality

```{r}
# Mortality analysis
```

The first analyses explore whether periods with higher vaccination coverage were accompanied by lower numbers of COVID-19 deaths.

## Vaccination and ICU Occupancy

```{r}
# ICU analysis
```

This section investigates the relationship between vaccination coverage and ICU occupancy over time.

# Current Progress

So far, we have

-   imported and cleaned the dataset,
-   explored the relevant variables,
-   produced initial visualizations,
-   started investigating the relationships between vaccination coverage, mortality, and ICU occupancy.

These analyses will be expanded and interpreted in greater detail in the final version of the website.

# Next Steps

For the final project we plan to

-   improve the visualizations,
-   perform additional descriptive analyses,
-   quantify the observed relationships where appropriate,
-   provide a comprehensive discussion of the findings and their limitations.
