---
title: "COVID-19 Vaccination and Disease Severity in Germany"
format: html
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(skimr)
library(infer)

covid_germany <- read_delim("data2/covid_germany.csv")
```

# Introduction

This project investigates the development of COVID-19 vaccination coverage in Germany and explores whether vaccination rates were associated with measures of disease severity during the pandemic.

We use publicly available data from the *Our World in Data* COVID-19 dataset, restricted to Germany.

# Dataset

The dataset contains daily observations for Germany between 2020 and 2024.

For our analyses we mainly use the following variables:

-   `date`
-   `people_fully_vaccinated_per_hundred`
-   `new_deaths_smoothed`
-   `icu_patients`
-   `new_cases_smoothed` (for context)

Before the analysis, we inspected the dataset for missing values, variable types, and implausible observations.

# Research Questions

Our project addresses the following questions:

1.  How did COVID-19 vaccination coverage in Germany change over time?

2.  Was higher vaccination coverage associated with lower COVID-19 severity?

    -   Was higher vaccination coverage associated with lower COVID-19 mortality?
    -   Was higher vaccination coverage associated with lower ICU occupancy?

# Preliminary Analysis

## Vaccination Coverage

### Research Question 1

#### **Model specification — RQ1**

-   **Response:** `people_fully_vaccinated_per_hundred`

-   **Predictor:** `date`

-   **Filter:** restricted to rows where `people_fully_vaccinated_per_hundred` is not missing (vaccination reporting only begins in early 2021)

-   **Method:** `geom_line` (raw trajectory)

-   **Expected output:** a time-series plot showing the trajectory of full vaccination coverage, from which we identify and qualitatively describe the main campaign phases (onset, steep rise, plateau near 78%)

-   **Interpretation:** the plot characterizes how full vaccination coverage evolved over time; no comparison to other variables or causal claims are made at this stage

#### Can the data support our RQ1? – Sanity Check

```{r}
glimpse(covid_germany)
```

We confirmed that `date`, `people_fully_vaccinated_per_hundred` and `new_vaccinations_smoothed`have the correct data types in order for us to create the planned line plot (Date and double) using glimpse().

```{r}
#check variable people_fully_vaccinated_per_hundred
covid_germany %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  summarise(start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  arrange(date) %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  summarise(max_gap = max(gap, na.rm = TRUE))
```

The Data coverage on vaccination starts in January of 2021 and ends in July of 2023. In this time range the data coverage is very good with a maximum gap of 1 day without coverage.

```{r}
#check variable new_vaccinations_smoothed
covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  summarise(start = min(date), end = max(date))

covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  arrange(date) %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  summarise(max_gap = max(gap, na.rm = TRUE))
```

By using the same method as before we now found out that the data for `new_vaccinations_smoothed`ranges from December 2020 until July 2023 datewise. Also just like with `people_fully_vaccinated_per_hundred` the coverage for this variable is very good.

As we explored in the *Domain-informed quality checks* part of our Proposal, the values of `people_fully_vaccinated_per_hundred` ranges from 0 to 78, which is enough variation to detect distinct phases over the available period.

```{r}
covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  summarise(
    min = min(new_vaccinations_smoothed),
    max = max(new_vaccinations_smoothed),
    mean = mean(new_vaccinations_smoothed),
    sd = sd(new_vaccinations_smoothed)
  )
```

The variation of values as well as the variance of `new_vaccinations_smoothed` is also going to be sufficient to answer our research question.

```{r}
covid_germany %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>% 
  ggplot(
    aes(x = date, y = people_fully_vaccinated_per_hundred)
  ) +
   geom_line(
      colour = "steelblue",
      linewidth = 1
    ) +
    labs(
      title = "COVID-19 vaccination coverage in Germany",
      subtitle = "People vaccinated per 100 inhabitants",
      x = "Date",
      y = "Vaccination coverage (%)"
    ) +
    scale_x_date(date_breaks = "2 months",
                date_labels = "%b\n%Y") +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 11),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10)
    )
```

Vaccination coverage increased rapidly after the campaign started in early 2021. During the first half of 2021 the slope is steep, indicating a fast rollout of vaccinations. Growth slowed substantially during late 2021 and eventually reached a plateau around 76–78% of the population. Afterwards only minor increases were observed until reporting ended in mid-2023.

```{r}
covid_germany %>%
  filter(!is.na(new_vaccinations_smoothed)) %>%
  ggplot(aes(date, new_vaccinations_smoothed)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Daily Vaccination Activity",
    subtitle = "7-day smoothed number of vaccinations",
    x = "Date",
    y = "New vaccinations"
  ) +
  theme_minimal()
```

In contrast to cumulative vaccination coverage, daily vaccination activity exhibits a pronounced peak during the first vaccination campaign in 2021. Afterwards vaccination activity declined steadily with only smaller temporary increases during later booster campaigns.

## Vaccination and Mortality

```{r}
# Mortality analysis
```

The first analyses explore whether periods with higher vaccination coverage were accompanied by lower numbers of COVID-19 deaths.

## Vaccination and ICU Occupancy

```{r}
# ICU analysis
```

This section investigates the relationship between vaccination coverage and ICU occupancy over time.

# Current Progress

So far, we have

-   imported and cleaned the dataset,
-   explored the relevant variables,
-   produced initial visualizations,
-   started investigating the relationships between vaccination coverage, mortality, and ICU occupancy.

These analyses will be expanded and interpreted in greater detail in the final version of the website.

# Next Steps

For the final project we plan to

-   improve the visualizations,
-   perform additional descriptive analyses,
-   quantify the observed relationships where appropriate,
-   provide a comprehensive discussion of the findings and their limitations.
